Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-11-25 rnni_convexity.tex
missing linebreaks + reference

The nearest neighbour interchange ($\nni$) graph, defined on the set of phylogenetic trees with adjacency relation given by the interchange operation of two sister clades, has been known in mathematical biology literature for nearly 50 years \autocite{Robinson1971-ql,Moore1973-kk}.
Considered with the metric given by the length of a shortest path (graph-distance), this graph becomes a metric space.
Its geometry has been extensively studied \autocite{Dasgupta2000-xa, Li1996-zw, Gordon2013-fw, De_Jong2016-al}.
An important property of the $\nni$ graph is that computing distance is NP-hard \autocite{Dasgupta2000-xa}.
As a result no algorithm exists to compute the distance in practical time.
A consequence is that tree search and sampling algorithms pose a significant challenge even for moderately sized trees
\todo{LC: Alex, is this the citation you wanted?}
\autocite{Whidden2018-tk}.

@@

\begin{figure}[H]
\centering
\includegraphics[width=.8\textwidth]{NNI_NP_proof}
\vspace{12pt}
\caption{Example of a shortest path between two trees $T$ and $R$ in $\nni$.
All leaves labelled with dots represent taxa in $\{1, \ldots, n\}$, squares represent taxa in $\{n+1, \ldots, 2n\}$.
In \autocite{Li1996-zw} it is proven that there is such a labelling that ensures that there is no shortest path in $\nni$ where the clusters (dots vs. squares) shared between $T$ and $R$ are preserved.
Instead, the depicted path where $n$ cherries are built and then sorted before being resolved is a shortest path}
\label{fig:NNI_NP_proof}
\end{figure}

@@

% \section{Ideas}
%
% [This section includes some ideas that we could possibly include in the paper]
%
% Using the same argument as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.
%
% \begin{proposition}
% If trees $T$ and $R$ and taxon $x$ are such that $|\rank_T(\parent_T(x)) - \rank_R(\parent_R(x))| = \delta$ then $d(T,R) \geq \delta$.
% \label{proposition:lower_bound_distance}
% \end{proposition}
%
%
% The paths computed by $\findpath$ depend on the order of input trees $T$ and $R$:
% $\findpath(T,R)$ does not necessarily produce the reverse path of $\findpath(R,T)$.
% An example can be found in Figure~\ref{fig:findpath_not_symmetric}.
% However, simulations suggest that the lengths of the two paths computed by $\findpath$ for one pair of trees are equal.
%
% \begin{conjecture}
% Let $T$ and $R$ be ranked trees.
% The lengths of paths $\findpath(T,R)$ and $\findpath(R,T)$ are equal.
% \end{conjecture}
%
% \begin{figure}[H]
% \centering
% \includegraphics[width=0.7\textwidth]{findpath_not_symmetric}
% \vspace{12pt}
% \caption{Two paths computed by $\findpath$. At the top $\findpath(T,R)$, at the bottom $\findpath(R,T)$}
% \label{fig:findpath_not_symmetric}
% \end{figure}
%
% % idea for description of max distance caterpillar trees:
% \begin{lemma}
% Let $T$ and $R$ be caterpillar trees.
% It is $d_c(T,R) = \Delta(\rnni)$ if, and only if, $T$ and $R$ share no induced triplet.
% \todo{define induced triplets}
% \end{lemma}
%
% This Lemma does only hold for caterpillar trees and not for trees with more than one cherry.
% Furthermore, triplets do not represent ranked trees uniquely as they do not contain information about the ranks of internal nodes (but they can uniquely define caterpillar trees as these only have one cherry)
%
% \begin{proof}
% %TODO
% % Insert the proof here
% Fix $T = (\ldots(1,2) \ldots ,n)$, use induction on $n$ and $\csort$ (taxon $n$ must be in cherry of $R$ if distance is max).
% \end{proof}
%
% According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given caterpillar tree $T$ is $(n-1)!$.
% The number of caterpillar trees with distance $\Delta(\rnni)$ from a given caterpillar tree is $2^{n-2}$.
%
%
% \subsection{Partition lattice}
%
% % Define Partition Lattice $\Pi_n$ and max chains in that lattice!
%
% \begin{theorem}
% The $\rnni$ graph on $n$ taxa is isomorphic to the graph of maximal chains of the partition lattice $\Pi_n$ where two maximal chains are connected by an edge if and only if they differ by exactly one partition.
% The corresponding metric spaces are isometric.
% \end{theorem}

>>>>

The nearest neighbour interchange ($\nni$) graph, defined on the set of phylogenetic trees with adjacency relation given by the interchange operation of two sister clades, has been known in mathematical biology literature for nearly 50 years \autocite{Robinson1971-ql,Moore1973-kk}.
Considered with the metric given by the length of a shortest path (graph-distance), this graph becomes a metric space.
Its geometry has been extensively studied \autocite{Dasgupta2000-xa, Li1996-zw, Gordon2013-fw, De_Jong2016-al}.
An important property of the $\nni$ graph is that computing distance is NP-hard \autocite{Dasgupta2000-xa}.
As a result no algorithm exists to compute the distance in practical time.
A consequence is that tree search and sampling algorithms pose a significant challenge even for moderately sized trees
\autocite{Whidden2016-kl}.

@@

\begin{figure}[H]
\centering
\includegraphics[width=.8\textwidth]{NNI_NP_proof}
\vspace{12pt}
\caption{Example of a shortest path between two trees $T$ and $R$ in $\nni$.
All leaves labelled with dots represent taxa in $\{1, \ldots, n\}$, squares represent taxa in $\{n+1, \ldots, 2n\}$.
In \autocite{Li1996-zw} it is proven that there is such a labelling that ensures that there is no shortest path in $\nni$ where the clusters (dots vs.\ squares) shared between $T$ and $R$ are preserved.
Instead, the depicted path where $n$ cherries are built and then sorted before being resolved is a shortest path}
\label{fig:NNI_NP_proof}
\end{figure}

@@

% \section{Ideas}
%
% [This section includes some ideas that we could possibly include in the paper]
%
% Using the same argument as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.
%
% \begin{proposition}
% If trees $T$ and $R$ and taxon $x$ are such that $|\rank_T(\parent_T(x)) - \rank_R(\parent_R(x))| = \delta$ then $d(T,R) \geq \delta$.
% \label{proposition:lower_bound_distance}
% \end{proposition}
%
%
% The paths computed by $\findpath$ depend on the order of input trees $T$ and $R$:
% $\findpath(T,R)$ does not necessarily produce the reverse path of $\findpath(R,T)$.
% An example can be found in Figure~\ref{fig:findpath_not_symmetric}.
% However, simulations suggest that the lengths of the two paths computed by $\findpath$ for one pair of trees are equal.
%
% \begin{conjecture}
% Let $T$ and $R$ be ranked trees.
% The lengths of paths $\findpath(T,R)$ and $\findpath(R,T)$ are equal.
% \end{conjecture}
%
% \begin{figure}[H]
% \centering
% \includegraphics[width=0.7\textwidth]{findpath_not_symmetric}
% \vspace{12pt}
% \caption{Two paths computed by $\findpath$.
% At the top $\findpath(T,R)$, at the bottom $\findpath(R,T)$}
% \label{fig:findpath_not_symmetric}
% \end{figure}
%
% % idea for description of max distance caterpillar trees:
% \begin{lemma}
% Let $T$ and $R$ be caterpillar trees.
% It is $d_c(T,R) = \Delta(\rnni)$ if, and only if, $T$ and $R$ share no induced triplet.
% \todo{define induced triplets}
% \end{lemma}
%
% This Lemma does only hold for caterpillar trees and not for trees with more than one cherry.
% Furthermore, triplets do not represent ranked trees uniquely as they do not contain information about the ranks of internal nodes (but they can uniquely define caterpillar trees as these only have one cherry)
%
% \begin{proof}
% %TODO
% % Insert the proof here
% Fix $T = (\ldots(1,2) \ldots ,n)$, use induction on $n$ and $\csort$ (taxon $n$ must be in cherry of $R$ if distance is max).
% \end{proof}
%
% According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given caterpillar tree $T$ is $(n-1)!$.
% The number of caterpillar trees with distance $\Delta(\rnni)$ from a given caterpillar tree is $2^{n-2}$.
%
%
% \subsection{Partition lattice}
%
% % Define Partition Lattice $\Pi_n$ and max chains in that lattice!
%
% \begin{theorem}
% The $\rnni$ graph on $n$ taxa is isomorphic to the graph of maximal chains of the partition lattice $\Pi_n$ where two maximal chains are connected by an edge if and only if they differ by exactly one partition.
% The corresponding metric spaces are isometric.
% \end{theorem}
